The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)

### 2018-06-07 analysis/paper/paper.Rmd
mca changes

#library(ScaledPiecesVB) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

@@

library(knitr)
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(ggpubr)
library(ggplot2)
library(ggrepel)
library(maptools)
library(ggmap)
library(devtools)
library(legendMap)
library(tab)

@@

Beyond the classification debate, the biggest issue with this type of tools comes from a functional standpoint, a problem that has also been quite debated [e.g. @binford_indian_1963; @de_la_pena_sobre_2011; @flood_moth_1980; @igreja_functional_2013; @leblanc_wedges_1992; @lucas_les_2004; @shott_bipolar_1999; @tixier_typologie_1963]. The issue surrounding this problem lies within the functional equifinality of these artifacts. Contrary to bipolar cores that are most of the times easily identified as cores, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard raw materials (e.g. bone, ivory, antler); and (2) as cores for the extraction of chips and small bladelets [@brantingham_early_2004]. Although largely debated, the ambiguity of this classification has, however, been ignored in some of the most recent literature, with some authors not acknowledging that bipolar evidence in stone tools might also result from other activities other than only reduction strategies [see e.g. @hiscock_dynamics_2015].

@@

In the case of the westernmost regions of Iberia, scaled pieces are commonly found in archaeological contexts ranging from the Upper Paleolithic to the Neolithic [e.g. @bicho_technological_2000; @carvalho_talhe_1998; @zilhao_o_1997]. While scarce, the majority of portuguese paleolithic studies [e.g. @almeida_terminal_2000; @goncalves_variabilite_2012] do not offer analytic interpretations for the presence of scaled pieces, simply using references that interpret them as bipolar cores for the extraction of small bladelets and chips to be supposedly used in composite tools [e.g. @zilhao_o_1997] or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009]. Zilhão [-@zilhao_o_1997] for example argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter. Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the portuguese early Neolithic.

While we agree that some of the elements might have been used as cores, these interpretations seem overly simplistic, and the authors seem to have little to no evidence to back their claims other than the inverse relationship mentioned above. This results in part from the lack of comprehensive studies regarding this type of artifacts at a regional scale. Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation and, thus, no direct association between these tools and the exploitation of hard-organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.

@@

The archaeological site of Vale Boi is located in the western edge of the Algarve region (southern Portugal) (Figure \@ref(fig:VBLocation)). The site can be found in a small valley following a river north-south for 2 km until it reaches the Atlantic Ocean. Archaeological deposits occupy an estimated area of over 10 000 sq. meters across a stepped slope marked at the top by a 10 meter-high limestone cliff face.

@@

Vale Boi’s lithic studies have revealed a general tendency for stable technological and functional patterns throughout the Upper Paleolithic [e.g. @bicho_ledge:_2012; @bradtmoller_lithic_2016; @cascalheira_cross-scale_2017; @cascalheira_influencia_2013; @cascalheira_vale_2012; @gibaja_funcion_2011; @marreiros_as_2009; @marreiros_lithic_2013; @marreiros_lithic_2015]. This can be explained, partially, by raw material availability. Most raw materials were procured locally, or regionally from deposits located at no more than 20 km away from the site [@bicho_paleolithique_2010]. 
Chert was the most used rock type for more complex retouched tools production, while quartz and greywacke were mostly used for flake extraction and simple retouched tools production. Other raw materials can be found at the site, but much more restricted, both diachronically and functionally, within each techno-complex. Schist, for instance, shows up in some occupations almost exclusively connected to mobile art. Dolerite was only  found in the Proto-Solutrean levels [@belmiro_o_2017], while chalcedony is mostly limited to the Proto-Solutrean and Solutrean levels [@cascalheira_vale_2012].

@@

## Reproducibility and open source materials

@@

Other characteristics of the assemblage are the low frequency of retouch (n = `r techno_data %>% filter(RetouchedBlank == "Yes") %>% summarise(n())`), of fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and of the presence of original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`). Still, when present, retouch is located in the lateral part of the artifacts, similar to what would define a side-scraper. In a very small number of cases, when striking platforms were detected, these are mostly flat. The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice. The large absence of platforms may, also, be the result of the use of the technological axis as the main functional axis. In fact, when identifiable, technological and typological axes coincide in `r inline_perc(techno_data, AxisCoincident, "Yes")`% of the cases. The longitudinal section shows a large variability of shapes, independent of raw material or chronology. Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of platform absence, might occur due to the functional stigmas and be dependent on the intensity of use for each artifact.

@@

```{r mestric_statistics, echo = FALSE}

@@

anova.test.area <- anova.test(Area ~ Chronology, techno_data)

anova.test.thickness <- anova.test(Thickness ~ Chronology, techno_data)

@@

As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use. Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straightforwardly performed. Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e. Length x Width) as an approximation for the overall dimensions of the artifacts. Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm. For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm. In general, mean values tend to be similar between raw materials, with some differences occurring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)). Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area ($F (`r anova.test.area[1]`, `r anova.test.area[2]`) = `r anova.test.area[3]`$, $p = `r anova.test.area[4]`)$ and Thickness ($F (`r anova.test.thickness[1]`, `r anova.test.thickness[2]`) = `r anova.test.thickness[3]`$, $p = `r anova.test.thickness[4]`)$.

@@

The patterns of use and rotation of damaged platforms seem to be the same across all chronologies. In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms. On the other end, a small number of chert and chalcedony artifacts (n= `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" & DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)). This seems to indicate different strategies of curation for coarse and fine grain raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity. However, when plotted against metric data, results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thickness of the pieces comprising each group. This seems to attest that the use of several axis in the same piece is not correlated with curation occuring in later phases of artifact use but instead to a probable difference in raw material performance.

@@

ggplot(n_axis, aes(x = factor(RawMaterial), y = perc, fill = factor(DamagedPlatforms))) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "RawMaterial", y = "%", fill = "Number of Damaged Platforms") +
  facet_wrap(~Chronology) +
  theme_bw() +
  theme(strip.text.x = element_text(size = 12))+
  scale_fill_manual(values=c("#00AFBB", "#E7B800", "#FC4E07", "#999999"))

@@

area_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Area", "", "Area (Length * Width)")
thickness_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Thickness", "Number of Damaged Platforms", "Thickness (mm)")

@@

@@

With few exceptions, most of the morpho-functional attributes show a fairly high degree of variability. Similarly to what was registered for the technological attributes most of the differences occur between raw materials rather than between chronologies. For this reason the results presented below focus only on the overall and between raw materials variability. Figure \@ref(fig:morphoattributes) shows the absolute frequencies of all qualitative variables recorded for each damaged platform. To simplify representation, and since we actually do not know which platform was the hammered one, opposed damage platforms are grouped into pairwise categories. When present, category 'Other' represents the cluster of attributes whose frequency was less than 10% within each variable.

In three (Scar Arrangement, Extension and Edge Delineation) of the six represented variables the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable.

For quartz artifacts the most common patterns are the combination of central/total scar distributions, both opposed platforms with angles wider than 45º, and a combination of either unifacial/unifacial or unifacial/bifacial scar distribution. Chert pieces, on the other hand, more typically present opposed platforms with damage occupying the whole platform width, and a weaker presence of unifacial/unifacial scar edge facial distribution. Still, with exception of Angle and Scar Facial Distribution, the overall trend for the morpho-functional variables is one of high variability, with a very large set of combinations appearing at very low frequencies.

To identify possible patterns of association between each one of the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis was run using the FactoMineR package. Results presented in Figure \@ref(fig:mca).

The first two dimensions of the MCA express 21.35% of the total dataset inertia, meaning that 21.35% of the individuals (or variables) cloud total variability is explained by the plane combining the first two dimensions. This is a rather small percentage but it is greater than the reference value that equals 18.92%. The variability explained by this plane is thus significant.

@@

opposed_plat_morpho()

>>>>

@@

library(knitr)
library(tidyverse)
library(broom)
library(FactoMineR)
library(factoextra)
library(ggpubr)
library(ggplot2)
library(ggrepel)
library(maptools)
library(ggmap)
library(devtools)
library(legendMap)
library(tab)
library(sjstats)
library(powerAnalysis)
library(epitools)
#library(ScaledPiecesVB)

@@

Beyond the classification debate, the biggest issue with this type of tools comes from a functional standpoint, a problem that has also been largely debated [e.g. @binford_indian_1963; @de_la_pena_sobre_2011; @flood_moth_1980; @igreja_functional_2013; @leblanc_wedges_1992; @lucas_les_2004; @shott_bipolar_1999; @tixier_typologie_1963]. The issue surrounding this problem lies within the functional equifinality of these artifacts. Contrary to bipolar cores that are most of the times easily identified as cores, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard raw materials (e.g. bone, ivory, antler); and (2) as cores for the extraction of chips and small bladelets [@brantingham_early_2004]. Although largely debated, the ambiguity of this classification has, however, been ignored in some of the most recent literature, with some authors not acknowledging that bipolar evidence in stone tools might also result from other activities other than only reduction strategies [see e.g. @hiscock_dynamics_2015].

@@

In the case of the westernmost regions of Iberia, scaled pieces are commonly found in archaeological contexts ranging from the Upper Paleolithic to the Neolithic [e.g. @bicho_technological_2000; @carvalho_talhe_1998; @zilhao_o_1997]. While scarce, the majority of portuguese paleolithic studies [e.g. @almeida_terminal_2000; @goncalves_variabilite_2012] have not presented, so far, any context-specific interpretations for the presence of scaled pieces. Most references broadly interpret these as bipolar cores for the extraction of small bladelets and chips to be  used in composite tools [e.g. @zilhao_o_1997] or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009]. Zilhão [-@zilhao_o_1997] for example argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter. Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the portuguese early Neolithic.

While we agree that some of the elements might have been used as cores, these interpretations seem overly simplistic, and the authors seem to have little to no evidence to back their claims other than the inverse relationship mentioned above. This results, in part, from the lack of comprehensive studies regarding this type of artifacts at a regional scale. Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation and, thus, no direct association between these tools and the exploitation of hard-organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.

@@

The archaeological site of Vale Boi is located in the western edge of the Algarve region (southern Portugal) (Figure \@ref(fig:VBLocation)). The site can be found in a small valley following a river north-south for 2 km until it reaches the Atlantic Ocean. Archaeological deposits occupy an estimated area of over 10 000 sq. meters across a stepped slope marked at the top by a 10 meter-high limestone cliff face [@bicho_ledge:_2012; @bicho_paleolithique_2010].

@@

Vale Boi’s lithic studies have revealed a general tendency for stable technological and functional patterns throughout the Upper Paleolithic [e.g. @bicho_ledge:_2012; @bradtmoller_lithic_2016; @cascalheira_cross-scale_2017; @cascalheira_influencia_2013; @cascalheira_vale_2012; @gibaja_funcion_2011; @marreiros_as_2009; @marreiros_lithic_2013; @marreiros_lithic_2015; @marreiros_lithic_2018]. This can be explained, partially, by raw material availability. Most raw materials were procured locally, or regionally from deposits located at no more than 20 km away from the site [@bicho_paleolithique_2010]. 
Chert was the most used rock type for more complex retouched tools production, while quartz and greywacke were mostly used for flake extraction and simple retouched tools production. Other raw materials can be found at the site, but much more restricted, both diachronically and functionally, within each techno-complex. Schist, for instance, shows up in some occupations almost exclusively connected to mobile art. Dolerite was only  found in the Proto-Solutrean levels [@belmiro_o_2017], while chalcedony is mostly limited to the Proto-Solutrean and Solutrean levels [@cascalheira_vale_2012].

@@

## Analysis, reproducibility and open source materials

@@

Other characteristics of the assemblage are the low frequency of retouch (n = `r techno_data %>% filter(RetouchedBlank == "Yes") %>% summarise(n())`), of fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and of the presence of original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`). Still, when present, retouch is located in the lateral part of the artifacts, similar to what would define a side-scraper. In a very small number of cases, when striking platforms are present, these are mostly flat. The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice. The large absence of platforms may, also, be the result of the use of the technological axis as main functional axis. In fact, when identifiable, technological and typological axes coincide in `r inline_perc(techno_data, AxisCoincident, "Yes")`% of the cases. The longitudinal sections show a large variability of shapes, independent of raw materials or chronologies. Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of platform absence, might occur due to the functional stigmas and be dependent on the intensity of use for each artifact.

@@

```{r metric_statistics, echo = FALSE}

@@

aov_area <- anova.test(techno_data, "Area")
cohens_area <- cohens.test(aov_area)
aov_area <- tidy(aov_area)

aov_thickness <- anova.test(techno_data, "Thickness")
cohens_thickness <- cohens.test(aov_thickness)
aov_thickness <-  tidy(aov_thickness)

@@

As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use. Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straightforwardly performed. Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e. Length x Width) as an approximation for the overall dimensions of the artifacts. Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm. For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm. In general, mean values tend to be similar between raw materials, with some differences occurring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)). Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (*F* (`r aov_area$df[1]`, `r aov_area$df[2]`) = `r aov_area$statistic[1]`, *p* = `r aov_area$p.value[1]`, *d* = `r cohens_area$cohens.f`), and Thickness (*F* (`r aov_thickness$df[1]`, `r aov_thickness$df[2]`) = `r aov_thickness$statistic[1]`, *p* = `r aov_thickness$p.value[1]`, *d* = `r cohens_thickness$cohens.f`).

@@

The patterns of use and rotation of damaged platforms seem to be the same across all chronologies. In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms. On the other end, a small number of chert and chalcedony artifacts (*N* = `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" | DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)). This seems to indicate different strategies of curation for coarse and fine grain raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity. However, when plotted against metric data, results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thickness of the pieces comprising each group. This seems to attest that the use of several axis in the same piece is not correlated with curation occuring in later phases of artifact use but instead to a probable difference in raw material performance.

@@

d_platforms <- ggbarplot(n_axis, x = "RawMaterial", y = "perc",
          color = "DamagedPlatforms", fill = "DamagedPlatforms",
          palette = c("#00AFBB", "#E7B800", "#FC4E07", "#999999"),
          facet.by = "Chronology", xlab = "", ylab = "%")

ggpar(d_platforms, legend = "right", legend.title = "Number of Damaged Platforms")

@@

area_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Area", "Number of Damaged Platforms", "Area (Length * Width)")
thickness_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Thickness", "Number of Damaged Platforms", "Thickness (mm)")

@@

\newline
```{r chisquare, echo = FALSE}

#Select and arrange morpho-functional variables for each pair of opposed platforms
morpho_data_opposed_platforms <- opposed_plat_morpho()

#Run chi-square and effect size test for each variable
chi_distribution <- CHI(morpho_data_opposed_platforms$ScarDistribution,morpho_data_opposed_platforms$RawMaterial)

chi_arrangement <- CHI(morpho_data_opposed_platforms$ScarArrangement,morpho_data_opposed_platforms$RawMaterial)

chi_extension <- CHI(morpho_data_opposed_platforms$ScarExtension,morpho_data_opposed_platforms$RawMaterial)

chi_delineation <- CHI(morpho_data_opposed_platforms$ScarEdgeDelineation,morpho_data_opposed_platforms$RawMaterial)

chi_faciality <- CHI(morpho_data_opposed_platforms$ScarFaciality,morpho_data_opposed_platforms$RawMaterial)

chi_angle <- CHI(morpho_data_opposed_platforms$Angle,morpho_data_opposed_platforms$RawMaterial)

@@

With few exceptions, most of the morpho-functional attributes show a fairly high degree of variability. Similarly to what was registered for the technological attributes, most of the differences occur between raw materials rather than between chronologies. For this reason the results presented below focus only on overall and between raw materials variability. Figure \@ref(fig:morphoattributes) shows the relative frequencies of all qualitative variables recorded for each damaged platform. To simplify representation, and to avoid wrong comparisons between hammered and active platforms, opposed damage platforms are grouped into pairwise categories. When present, category 'Other' represents the cluster of attributes which frequency was less than 10% within each variable.

In three (Scar Arrangement, Extension and Edge Delineation) of the six represented variables the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable. A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = `r round(chi_arrangement$chi_test$p.value, 3)`, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`). The effect size statistic (Phi) suggests, however, a low practical significance for Scar Arrangement and a large pratical significance for Scar Faciality.

For quartz artifacts the most common patterns are the combination of central/total scar distributions, both platforms with angles wider than 45º, and a combination of either unifacial/unifacial or unifacial/bifacial scar distribution. Chert pieces, on the other hand, more typically present opposed platforms with damage occupying the whole platform width, and a weaker presence of unifacial/unifacial scar edge facial distribution. Still, with exception of Angle and Scar Facial Distribution, the overall trend for the morpho-functional variables is one of high variability, with a very large set of combinations appearing at very low frequencies.

@@

#PLot categorical morpho-function variables

plot_morpho_var(morpho_data_opposed_platforms)

To identify possible patterns of association among the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis was performed. Results are presented in Figures \@ref(fig:mca). In this analysis raw material was used as a qualitative supplementary variable.

The first two dimensions of the MCA express 20.92% of the total dataset inertia, meaning that only that percentage of the total variability is explained by the plane combining the first two dimensions. An inspection to the screeplot presented in Figure \@ref(fir:screeplot) suggests to restrict the analysis to the description of the first 4 dimensions. These dimensions present an amount of inertia slightly greater than that obtained by the 0.95-quantile of random distributions (37.74% against 34.37%). Still, it is a rather small percentage, somehow attesting the high variability suggested by our interpretation of Figure \@ref(fig:morphoattributes), and suggesting that patterns identified by previous studies [e.g. @de_la_pena_qualitative_2015; @de_la_pena_sobre_2011], in which certain combination of attributes were used to discriminate scaled pieces functionalities, are difficult to apply to the assemblage used in this study.

The high variability of the assemblage is also well noticeable in the biplots presented in Figure \@ref(fig:biplot).

```{r mca, echo = FALSE}

# Multiple Correspondence Analysis

mca.rawmaterial()

```{r screeplot, echo = FALSE}

# Screeplot
fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "#00AFBB", barcolor = "#00AFBB")

```{r biplot, echo = FALSE}

# Biplot for 4 main MCA dimensions

fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "#00AFBB", barcolor = "#00AFBB")
